---
title: "Assignment2 Whanganui Property Rates"
author: "John Lavack 20001249"
date: "2026-08-09"
output: html_document
---

```{r attach libs, echo = FALSE, warning = FALSE, message = FALSE}
library(tidygraph)
library(tidyverse)
library(readxl)
library(ggdist)
library(maps)    
library(sf)  
library(patchwork)

```

```{r read in data, message = FALSE}
#data in Draft Rates by property 26.27 - statistical purposes 1 in two sheets. Data and Unit Rates

#===== Read in data ======
Data <- read_excel("Draft Rates by property 26.27 - statistical purposes 1.xlsx", sheet = "Data", skip = 2) %>%
  drop_na(property_no) #first two rows are blank, Can map headers from unit_rates reference data set. Last rows are totals, but totals have no property number, so drop. 
# Unit_Rates <- read_excel("Draft Rates by property 26.27 - statistical purposes 1.xlsx", sheet = "Unit Rates")
wdc_rates <- read_csv("wdc_rates_qv.csv") %>% 
  mutate(rate_yr = factor(rate_yr, levels = c("22_23", "23_24", "24_25", "25_26", "26_27"), ordered = TRUE))

#CPI data for indexing
CPI <- read.csv("CPI.csv") %>%
  mutate(Date = as.Date(Date, format = "%d-%m-%Y"),
         year = year(Date)) %>%
  group_by(year) %>%
  summarise(CPI = mean(CPI..Annual...change.))

#===== map setup begin ======
nz <- read_sf("https://www.massey.ac.nz/~jcmarsha/data/nz_map/sa3_data.sqlite")

ta_nz <- nz %>% 
  group_by(territorial_authority, regional_council) %>%
  summarise(geometry = st_union(GEOMETRY), .groups = 'drop')

whanganui <- nz %>% filter(territorial_authority == "Whanganui District")
bbox <- st_bbox(whanganui)

w_bbox <- list(x = c(1767500, 1779000), y=c(5574000, 5585000))
#===== map setup finish ======

#===== Mapping Suburbs begin =====
St_John_Streets <- c('Acton Pl', 'Aiken Rd', 'Alexa Pl', "Brassey Rd","Bristow St", "Chittick Pl","Collier Pl", "Dickson Cres", "Halswell St","Halswell Street","Hillside Tce", "Kent Rd","Linda Pl", "Ngapuhi St", "Oakland Ave","Parkes Ave", "Parsons St", "Porritt St", "Rawhiti Pl", "Smart Tce", "St Leonard St","Taranaki St","Tarata St", "Tregarth St","Tulloch St", "Virginia Rd", 'Peakes Rd', 'Victoria Park', 'Virginia Lake (Rotokawau)')

Otamatea_Streets <- c('Sewerage Pump Turere', 'Babbage Pl', 'Beachcroft', "Chelsea Cl", "Cotswolds Cl", "Diana Pl", "Eaton Cres", "Edith Collier Dr", "Edmonds Dr", "Goldsbury Pl", "Gumtree Rise", "Holdsworth Dr", "Kawau Pl", "Lancewood Pl", "Lithgow Dr", "Mannington Rd", "Montgomery Rd", "Norfolk Dr", "Otamatea Reserve", "Perrett Dr", "Pickwick Rd", "Rotokawau Dr", "Sandcroft Dr", "Sandy Lane", "St Johns Hts", "Suffolk Dr", "Tasman Views", "Tirimoana Pl", "Turere Pl","Virginia Hts", "Webster Pl", "Pukeko Drive", 'Tangaroa View', 'Middens Way')

Springvale_streets <-c('Addenbrooke Tce')

Whanganui_Centre_Streets <- c('Pitt St', 'Peat Pl', 'Peat St')

#===== Mapping suburbs finish =====

#===== Data wrangling begin =====
streets <- wdc_rates %>% 
  filter(suburb == "St Johns/Otamatea") %>% 
  select(property_no, rate_yr, property_address, suburb) %>%
  mutate(
    house_num = as.integer(str_extract(property_address, "^\\s*(\\d+)")), 
    street = str_remove(property_address, "~.*"),
    street = str_remove(street, "-."),
    street = str_remove(street, "^(?i:unit\\s+)?(?:[\\d/][\\d/A-Za-z]*\\s+)+"), #handles numbers and numbers like 9A,alos UNIT 6C / 59 and
  ) %>%
  mutate(
    suburb = case_when(
      #house numbers on great north road. great north road spans both. Lower number in St Johns up to 117. Then 119 and 121 are St Johns. 118, 120 and 122 onwards are otamtea
      street == "Great North Rd" & is.na(house_num) ~ "St Johns Hill",
      street == "Great North Rd" & house_num <= 117 ~ "St Johns Hill",
      street == "Great North Rd" & house_num %in% c(119, 121) ~ "St Johns Hill",
      street == "Great North Rd" & house_num >= 118 ~ "Otamatea (Whanganui District)",
      #easier ones
      street %in% St_John_Streets ~ "St Johns Hill",
      street %in% Otamatea_Streets ~ "Otamatea (Whanganui District)",
      street %in% Springvale_streets ~ "Springvale",
      street %in% Whanganui_Centre_Streets ~ "Whanganui Centre",
      TRUE ~ suburb
    )
  ) %>% select(property_no, rate_yr, suburb) #for left join

#join tdatasets to get what I need. Remove nulls from data set.
Rates <- wdc_rates %>% 
  filter(!is.na(assessment_number)) %>%
  left_join(
    Data %>% select(property_no, SUIP), 
    by = 'property_no') %>%
  mutate( #mapping qv_category to res/com/farm for simpler groupings
    Category = case_when(
      `qv category` %in% c('Commercial', 'Industrial') ~ 'Commercial', 
      `qv category` %in% c('Farming', 'Dairy', 'Pastoral', 'Forestry', 'Horticulture', 'Mining') ~ 'Farming',
      `qv category` %in% c('Other', 'Specialist', 'Utility asset') ~ 'Specialist',
      `qv category` %in% c('Residential', 'Lifestyle') ~ 'Residential',
      TRUE ~ `qv category`),
    SUIP = case_when(
      `qv category` == 'Mining' ~ 1,
      `qv category` == 'Residential' & is.na(SUIP) ~ 1,
      TRUE ~ SUIP)
    ) %>%
  left_join(streets, by = c('property_no', 'rate_yr'), suffix = c("", "_new")) %>%
  mutate(suburb = coalesce(suburb_new, suburb)) %>%
  mutate(
    suburb = recode(suburb,
    "Bastia/Durie Hills Putiki" = "Bastia-Durie Hill-Putiki",
    "Whanganui Central" = "Whanganui Centre",
    "Whanganui Rural" = "Whanganui District Rural",)
  ) %>%
  mutate(year = case_when(
    rate_yr == '22_23' ~ 2022,
    rate_yr == '23_24' ~ 2023,
    rate_yr == '24_25' ~ 2024,
    rate_yr == '25_26' ~ 2025,
    rate_yr == '26_27' ~ 2026,
    TRUE ~ NA
  )) %>%
  mutate(Category = factor(Category, 
                           levels = c('Specialist', 'Farming', 'Commercial', 'Residential'))) %>% #for consistent color coding
  select(-suburb_new) 
#===== Data wrangling finish =====

#===== Colour palette setup =====
whanganui_palette <- c( #from inspecting the page elements of the Whanganui website, then color dropper picking from the coat of arms
  "#2757A2", 
  "#AD1619", 
  "#FEC30E", 
  "#008242", 
  "#DCC0E2", 
  "#C7E9F3",
  "#004925",
  "#D4582A",
  "#372502",
  "#8B514E",
  "#93735B", 
  "#687928",
  "#FEEAD7")

# font is Inter sans-serif


#===== additional config =====
x_labels <- c('2022/23', '2023/24', '2024/25', '2025/26', '2026/27')

```

```{r}
# sum of {CBD rates, UAGC, General Rates, Water Rates, Stormwater, Firefighting capacity, Roading, Wastewater, Earthquake, Kerbside, Forestry, Roading Resilience, Trade Waste} = total
# AND sum of {CBD, CBDA, CBD, ESC, ESF, ESR, GC1, GCS, GFA10, GFA3, GFA4, GFA5, GR1, GRA1, GRA2, GRA5, GRS, KS, MAW, MAWD, MAWR, RDA, RDAS, RDB, RDBS, RDF, RRC, RRF, RRR, SD, SDO, SDOS, SMO, SMOH, SRWV1, SRWV2, UAGC, WC, WCM, WFC, WS, WSH, WSRE, WWS, FE, TWCOD, TWFlow, TWTSS, TWWWTP} = total. 
# first set of sums have conditional rules that look along row 2 (where unit rates names are above the columns in set 2), checks which match the column headers in the first set, then sums the columns. It is a classification hierarchy grouping process. So, column BQ is for CBD rates, the sum looks for all columns in the group `CBD Rates` then sums only those. Of which there are three: CBD, CBDA and CBDB. Same deal for all the others. 

# The requirement is a 2 page flyer. Therefore, expect 2- 4 graphics. Try 2 simpler ones and 2 more complex ones.  
# 1. Which areas have the highest rates over time? OR which areas have the highest change this time?: a map of whanganui with heat colors. 
# Can show areas with QV average for area and rates average per area to eyeball if higher = higher or not.   
# 2. What contributes the most to rates increases this year, using data and the unit rates to look at the contributing components of each rate element.
#    find the max column, then reshape to get property no, value, contirbuting rate code, then the hierarchy it belongs to. And % contribution to rate increase. Can also look at the aggregate part too... could be fun. 
# 3. Look at rates variances 
# 4 land area changes or land usage changes and impact on rates. Perhaps I can color in the map or look at suburbs and their split of qv_categories vs rates changes over time. i.e. increase in commerical and larger increase in rates? Will play with that. 
# Lead with a gimme: how rates have changed over time in whanganui. the year on year % change overall. As a lead in. 
# Fan chart 
```


Here's the plan:
1. look at rates by category then highlight that residential rates are what most people think of when they think rates. Show bar chart with nubmer of members by category and variance of rates by category. Bar chart with box chart, or the box like chart that shows all the points. 
2. Look at residential rates by SUIP. This is going to be done with a half eye chart. 
3. Show residential rates through time split by multiple unit or single unit SUIP and compare to average annual inflation (sum up year inflation and divide by 4, get this from RBNZ) This is spaghetti chart and ribbon chart side by side. 
4. Show suburbs rates on map. side by side with a heat map of suburb rates through the years

Maybe also look at how rates varied year on year within categories.
Rain cloud for ggdits.
use line ribon for year on year variation

# Graph 0: rates over time vs RBNZ inflation
```{r graph 0, fig.width = 10, fig.height=5}
#RBNZ data I just copied from their website. Didn't seem like worth all the effort to download the full dataset.Rate yaers frun from start july to end June (source: my ever increasing rates bill :( )
#NZ financial year is from start march to end april, so I need to get the NZ rates year quarter inflation values to map the rates to inflation correctly

#RBNZ pink #ed1166
RBNZ_inflation <- CPI %>% filter(year >= 2022)
CPI_2022 <- RBNZ_inflation %>% filter(year == 2022) %>% pull(CPI)

#indexing rates to inflations, then getting % change ot compare next to CPI
timeseries_data <- Rates %>% #filter(Category == 'Residential') %>%
  arrange(property_no, year) %>%
  left_join(RBNZ_inflation, by = "year") %>%
  mutate(indexed_rates = (`total rates` / CPI) * CPI_2022) %>%
  group_by(property_no) %>%
  mutate(pct_change = (indexed_rates - lag(indexed_rates)) / lag(indexed_rates) * 100,
         pct_change = replace_na(pct_change, 0)) %>%
  ungroup() %>%
  select(property_no, year, `total rates`, indexed_rates, pct_change, Category)

G0_data <- timeseries_data %>%
  group_by(Category, year) %>%
  summarise(mean_pct_change = mean(pct_change), .groups = 'drop') 

G0_data %>%
  ggplot(aes(y = mean_pct_change, x = year, col = Category)) +
  geom_line() +
  geom_point(size = 2,show.legend = FALSE) +
  geom_line(data = RBNZ_inflation, aes(y = CPI, x = year, color = "RBNZ Inflation Rate"), linewidth = 1, linetype= 'dashed') +
  geom_point(data = RBNZ_inflation, aes(y = CPI, x = year, color = "RBNZ Inflation Rate"), size = 2, show.legend = FALSE) +
  #Scale arguments
  scale_x_continuous(labels = x_labels) +
  scale_y_continuous(limits = c(round(min(G0_data$mean_pct_change),0), 
                                  round(max(G0_data$mean_pct_change),0)),
                      breaks = scales::breaks_width(10)) +
  scale_color_manual(values = c(
    setNames(whanganui_palette[1:4], c('Specialist', 'Farming', 'Commercial', 'Residential')),
    "RBNZ Inflation Rate" = "#ed1166"),
    name= "Legend") +
  #labels
  labs(title = 'Whanganui year on year mean rate percentage change against RBNZ CPI',
       subtitle = 'Rates changes are usually higher than CPI, except for 2026/27',
       x = '',
       y ='%',
       color = 'Legend',
       caption = 'Rate values have been index linked to 2022 inflation in order to compare like.' ) +
  #theme
  theme_minimal() +
  theme(legend.position = 'top',
        panel.grid.minor.x = element_blank())

```


#Graph 1: rates by category

```{r graph 1, fig.width=15, fig.height=7}
#===== Column chart =====#
Column <- Rates %>%
  filter(rate_yr == '26_27') %>% #avoid double counting
  group_by(Category) %>%
  summarise(Count = n()) %>%
  # mutate(Category = factor(Category, levels = c('Specialist', 'Farming', 'Commercial', 'Residential'))) %>%
  ggplot(aes(x=Count, y = Category, fill= Category)) +
  geom_col() +
  #scales
  scale_fill_manual(values = whanganui_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) + #stop the clipping off of the x axis values
  #labels
  labs(title = 'Rateable units by category',
       subtitle = '88.4% of ratable units are residential',
       y = '', 
       x = 'Ratable units') +
  #theme
  theme_minimal() +
  theme(plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = 'none',
        panel.grid.major.y = element_blank()) 

#===== box chart =====#
Box <- Rates %>%
  filter(rate_yr == '26_27') %>% #avoid double counting
  # mutate(Category = factor(Category, levels = c('Specialist', 'Farming', 'Commercial', 'Residential'))) %>%
  ggplot(aes(x=log(`total rates`), y = Category, col= Category)) +
  geom_point(position = position_jitterdodge(jitter.width = 1), alpha = 0.3, size = 1) +
  geom_boxplot(col = 'black', fill = NA, outlier.shape = NA, size = 0.7) +
  #scales
  scale_color_manual(values = whanganui_palette) +
  #labels
  labs(title = 'Rates distribution by category',
       subtitle = 'Residential rates varied the least compared to other categories',
       y = '', 
       x = 'Log of total rates') +
  #theme
  theme_minimal() +
  theme(plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 12),
        legend.position = 'none',
        panel.grid.major.y = element_blank(),
        axis.text.y = element_blank())

#===== patchwork the two together =====#
(Column | Box) +
  plot_annotation(title = "Rates by Category in year rating 2026/27",
                  theme = theme(plot.title = element_text(size = 30)))
```

#Graph 2: Residential rates change  through time. % change

```{r graphn 2, fig.height=8, fig.width=10}
#===== ribbon chart =====#
#redline shows median change
ribbon_chart <- timeseries_data %>%
  group_by(year, Category) %>%
  median_qi(pct_change, .width = c(0.25, 0.5, 0.75, 0.95)) %>%
  ggplot(aes(x = year, y = pct_change, ymin = .lower, ymax = .upper, color = Category)) +
  geom_lineribbon(color = "#DC0100", alpha = 0.5, linewidth = 0.7) + #from the vignette https://mjskay.github.io/ggdist/articles/lineribbon.html
  geom_vline(aes(xintercept = 2023), linetype ='dashed') + 
  facet_wrap(~ Category, scales = "free_y") +
  #scales
  scale_fill_manual(values = c("#C7E9F3", "#92B8D8", "#5C88BD","#2757A2"),
                    name = 'Quantile') +
  scale_x_continuous(labels = x_labels) +
  #labels
  labs(title = 'Inflation indexed residential rates change year on year',
     subtitle = 'Rate % change varied the most in the 2023/24 year, when the 2022 QV revaluation was applied.\nResidential rates have varied the least.',
     y = 'Percentage change', 
     x = '') +
  #Theme
  theme_minimal() +
  theme(legend.position = 'top',
        panel.grid.minor.x = element_blank(),
        strip.text = element_text(face = "bold"))

#===== Cumulative percent change =====#
cumulative_data <- timeseries_data %>%
  group_by(year, Category) %>%
  summarise(pct_change = mean(pct_change), .groups = 'drop') %>%
  group_by(Category) %>%
  arrange(year) %>% #should be arranged already... paranoid
  mutate(cum_chg = cumsum(pct_change))

avg_cum_perc <- cumulative_data %>% group_by(year) %>% summarise(avg = mean(cum_chg))

cumulative_chart <- cumulative_data %>%
  ggplot(aes(x = year, col = Category)) +
  geom_line(aes(y = cum_chg), linewidth = 0.8) +
  geom_line(data = avg_cum_perc, aes(x = year, y = avg, col = 'Average'),
            linewidth = 1, linetype = 'dashed') +
  #scales
  scale_x_continuous(labels = x_labels) +
  scale_y_continuous(limits = c(round(min(cumulative_data$cum_chg),0), 
                                round(max(cumulative_data$cum_chg),0)),
                    breaks = scales::breaks_width(10)) +
  scale_color_manual(values = c(
      setNames(whanganui_palette[1:4], c('Specialist', 'Farming', 'Commercial', 'Residential')),
      "Average" = "black")) +
  #labels
  labs(
    title = 'Cumulatively, residential rates have changed the least',
    subtitle = "Consistently just below the average",
    x = '',
    y = 'Percent'
  ) +
  #theme
  theme_minimal() +
  theme(legend.position = 'top',
      strip.text = element_text(face = "bold"))

#patchwork output together
(ribbon_chart / cumulative_chart)
  
  
```

Residential rates change the most in terms of median % change. 

#Graph 3: SUIP Residential
do new builds have higher rates? No, it doens't appear so. 

```{r}
year_built <- Rates %>% 
  group_by(property_no) %>%
  summarise(Count= n()) %>% 
  mutate(Year_built = case_when(
    Count == 1 ~ 2025,
    Count == 2 ~ 2024, 
    Count == 3 ~ 2023, 
    Count == 4 ~ 2022,
    TRUE ~ 2021
  )) %>% select(-Count)

Rates %>% left_join(year_built, by = 'property_no') %>%
  group_by(Year_built) %>%
  summarise(mr = median(`total rates`))
  
```




# Graph 3 - Map

```{r graph 4 map, fig.width=15, fig.height=8}
#===== map code =====
graph_3_data <- Rates %>% 
  filter(year == 2026, Category == 'Residential') %>%
  group_by(suburb) %>%
  summarise(median_rates = median(`total rates`),
            count = n(), 
            .groups = 'drop')

suburb_map <- whanganui %>%
  left_join(graph_3_data, by = c('sa3_name' = 'suburb')) %>%
  ggplot() +
  geom_sf(mapping=aes(fill = median_rates), col = 'white', linewidth = 1, alpha = 0.5) +
  coord_sf(xlim = w_bbox$x, ylim = w_bbox$y) + #need to change the rural district label to be visible
  geom_sf_label(mapping=aes(label=sa3_name), size=3) +
  geom_sf_text(mapping = aes(label = as.character(count)), size = 2.5, vjust = 3) + #count of ratable units
  annotate("label", x = 1777000, y = 5584000,
           label = "Whanganui District Rural", size = 3, fill = 'white') + #play with this to get the piosition right
  scale_fill_gradientn(colors = c("#008242", "#FEC30E", "#AD1619"),
                       limits = c(min(graph_3_data$median_rates), max(graph_3_data$median_rates))) +
  labs(title = 'Rates by suburb in year 2026/27',
       subtitle = '',
       x = '',
       y = '') +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

#===== heatmap code =====
heatmap_data <- Rates %>% 
  filter(Category == 'Residential') %>%
  select(`total rates`, suburb, year) %>%
  group_by(year, suburb) %>%
  summarise(median_rates = median(`total rates`),
             .groups = 'drop')

suburb_heatmap <- heatmap_data %>%
  ggplot(aes(y = suburb, x = year, fill = median_rates)) +
  geom_tile() +
  geom_text(aes(label = round(median_rates,0)), size = 3.5, na.rm = TRUE) +
  scale_fill_gradientn(colors = c("#008242", "#FEC30E", "#AD1619"),
                     limits = c(min(heatmap_data$median_rates), max(heatmap_data$median_rates))) +
  scale_x_continuous(breaks = c(2022, 2023, 2024, 2025, 2026),, labels = x_labels) +
  labs(title = 'Median Rates by Suburb from 2022/23 to 2026/27',
       subtitle = '',
       x = '',
       y = '') +
  theme_minimal()

#patchwork the plot together.
(suburb_map | suburb_heatmap) +
  plot_annotation(title = 'Median residential rates can vary more than $1,000 by suburb',
                  subtitle = 'Otamatea has the highest residential rates in Whanganui, despite having the fewest rateable units')
```

#Who pays the most?

```{r}
# Total of rates paid is most in gonville
total_rates <- sum(Rates$`total rates`)

Rates %>% 
  filter(Category == "Residential") %>%
  group_by(year, suburb) %>%
  summarise(total_rates_paid = sum(`total rates`),
            units = n(),
            .groups = "drop") %>%
  mutate(prop_rates = total_rates_paid / sum(total_rates_paid),
         average_rates = total_rates_paid / units) %>%
  arrange(desc(year), desc(prop_rates))

#per house, the proportional rates are hghest in omatea
Rates %>% 
  filter(Category == "Residential") %>%
  group_by(year) %>%
  mutate(prop = 100 * `total rates` / sum(`total rates`)) %>%
  ungroup() %>%
  group_by(year, suburb) %>%
  summarise(mean_prop = mean(prop), 
            sum_prop = sum(prop),
            .groups = "drop") %>%
  arrange(year, desc(mean_prop))



timeseries_data %>% 
  left_join(Rates %>% select(property_no, suburb), by = "property_no") %>%
  group_by(year, suburb) %>%
  summarise(total_rates_paid = sum(`total rates`))
```



#Graph 4 - rain clouds by SUIP

Are the rich contributing? Look at total rates paid by residential suburb
Maybe look at the stat half eye by subrub actual rates?
what about the SUIPs.


```{r, fig.height=10, fig.width=10}
# for one graph I will show, with the density eye plot the categorical density of rates by qv category. As they vary heaps
# lots of categories, maybe too many. I'll try mapping them ot the Data res/com/faarm ones ot get a clearer graph
widths = c(0.5, 0.8, 0.95)

#identify outliers by category, then also plot them on the graph., Use a simple IQR method

Rates %>% 
  filter(Category == 'Residential', !is.na(SUIP)) %>%
  mutate(SUIP_grp = case_when(
    SUIP == 0 ~ '0',
    SUIP == 1 ~ '1',
    SUIP == 2 ~ '2', 
    SUIP %in% c(3,4,5) ~ '3 - 5',
    SUIP %in% c(6,7,8,9,10) ~ '6 - 10',
    SUIP >10 ~ '11+'
    ),
    SUIP_grp = factor(SUIP_grp, levels = c('0', '1', '2', '3 - 5', '6 - 10', '11+'))) %>%
  # mutate(outlier = `total rates` < quantile(`total rates`, 0.25) - 1.5 * IQR(`total rates`) |
  #          `total rates` > quantile(`total rates`, 0.75) + 1.5 * IQR(`total rates`)) %>%
  # filter(!outlier) %>%
  ggplot(aes(x = `total rates`, y = SUIP_grp, fill = SUIP_grp)) +
  # stat_interval(
  #   mapping = aes(fill_ramp = after_stat(level)),
  #   .width = widths) +
  stat_halfeye(
    mapping = aes(fill_ramp = after_stat(level)),
    fill_type = "segments",
    .width = widths,
    alpha = 0.5,
    show.legend = FALSE,
    col = 'black',
  ) +
  stat_dotsinterval(
    mapping = aes(fill_ramp = after_stat(level)),
    .width = widths,
    side = "bottom",
    layout = 'weave',
    scale = 0.5,
    slab_linewidth = NA,
    show.legend = FALSE,
    quantiles = 100) +

  # stat_summary(geom = "point", fun = median, col = 'black', position = position_nudge(y = -0.05)) +
  scale_fill_ramp_discrete(na.translate = FALSE) +
  theme_minimal() 
  # scale_color_manual(values = c("#FEC30E","#6190F7", "#69A241"), aesthetics = 'slab_fill')

#copied code from assignment 1
# # +
#   geom_vline(xintercept = mean_value, col = 'black', lty = 'dashed') +
#   #add an annotation over hte y axis ot indicate the number of sites
#   annotate("text", 
#            x = 0, # Position x near the counts (right side)
#            y = y_levels[5], # Position y at the top-most category level
#            label = "(Sites, Samples)", 
#            size = 3, 
#            hjust = 1.2, # Left align the header with the counts
#            vjust = -3, # Push it up above the label
#            family = "mono") +
#   #group mean annotation
#   annotate("text", 
#            x = 2, # Position x near the counts (right side)
#            y = y_levels[5], # Position y at the top-most category level
#            label = paste0('pooled mean = ', round(mean_value, 2)), 
#            size = 3, 
#            hjust = -0.1, # keep it to the right of the mean line
#            vjust = -7, # Push it up over the geom
#            family = "mono") +
#   scale_y_discrete(labels = label_lookup) + #applying the custom labels 
#   #colors go outside in
#   scale_color_manual(values = c("#f9e4c4","#79b9e7", "#4277af")) + 
#   scale_x_continuous(breaks= seq(0, max(plot_data$`Clarity (BDISC)`), by = 1)) +
#   coord_cartesian(xlim = c(0, 12), clip = "off") +
#   labs(
#     title = 'Is oversampling bias making rivers less clear?',
#     subtitle = 'No, but increased activity around rivers does',
#     caption = "Pasture land and urban areas have mean water clarity below the pooled mean, despite high or low sample sites \n while native forests are above it. There appears to be no trend in altitude. \n Sampled sites are all between Wellington and Palmerston in NZ",
#     x = 'Water clarity (Meters of depth visibility)', 
#     y = '') +
#   theme_lawa() 
```
```{r}
Rates %>%  
  mutate(SUIP_grp = case_when(
    is.na(SUIP) ~ 'Unknown',
    SUIP == 0 ~ '0',
    SUIP == 1 ~ '1',
    SUIP == 2 ~ '2', 
    SUIP %in% c(3,4,5) ~ '3 - 5',
    SUIP %in% c(6,7,8,9,10) ~ '6 - 10',
    SUIP >10 ~ '11+',
    TRUE ~ as.character(SUIP)
  )) %>%
  group_by(SUIP_grp) %>%
  summarise(count = n())
```

=============================================================================

